I can see clearly now: reinterpreting statistical significance

### 2018-10-08 doc.tex
a few more quick comments

\textbf{Abstract}: Null hypothesis significance testing remains popular despite decades of concern about misuse and misinterpretation. We believe that a significant part of the problem is due simply to language: significance testing has little to do with other meanings of the word ``significance''. Despite the limitations of null-hypothesis tests, we argue here that they remain useful in many contexts as a guide to whether a certain effect can be seen \emph{clearly} in that context (e.g. whether we can clearly see that a correlation or between-group difference is positive or negative). We therefore suggest that \pvals resulting from null-hypothesis tests be described using the language of statistical \emph{clarity} rather than statistical \emph{significance}. This simple semantic change could substantially enhance clarity in statistical communication.

@@

Statisticians and scientists have bemoaned the shortcomings of null hypothesis significance testing (NHST) for nearly a century \citep{Cohen1994}.
Books and articles proposing the de-emphasis or abandonment of the \pval have been cited thousands of times \citep{Cohen1994, Goodman1999, Wilkinson1999, ZiliakandMcCloskey2008, WassersteinandLazar2016}. These works plead for a focus on effect sizes and confidence intervals, and point out that a zero, ``null effect'' is unrealistic or impossible in most fields outside of the hard physical sciences \citep{Meehl1990, Tukey1991, Cohen1994}. Yet, \pvals without confidence intervals (or even effect sizes) and references to null effects still pervade the scientific literature at all levels up to and including high-impact journals.

In a meta-analysis of 356 studies \citet{Bernardietal.2017} found that 72\% of studies contained an ambiguous use of the term ``significant'', 49\% interpreted non-significant effects as zero effects, and 44\% failed to report a comprehensible effect size. The misuse and misinterpretation of NHST is so frequent that there has been a recent push for drastically reducing \citep{SzucsandIoannidis2017} or abandoning \citep{McShaneetal.2017} its use. Other prescriptions have included calls for the complete abandonment of frequentist statistics \bmb{CITE: is it OK to include this here, or it too distracting?}, or for the use of a more strenuous significance threshold (e.g. $p < 0.005$: \citealt{Benjaminetal.2018}); however, the former seems impractical, while the latter is unlikely to reduce the misuse and misinterpretation of \pvals, or the publication bias imposed by any \pval threshold \citep{Ridleyetal.2007}.

\citet{Kramer2011} points out that \citep{ZiliakandMcCloskey2008} ask how could NHST possibly have survived ``\ldots if it is as idiotic as we and other critics have so long believed''. We believe that NHST itself is not as idiotic as suggested; the idiocy associated with NHST is the language we use in describing it. A strong argument in favor of NHST is that testing the null hypothesis (even though we often know that it's false, and can never prove that it's true) is a proxy for asking whether we can see a difference clearly. In many cases, this comes down simply to whether we can be confident of the \emph{sign} of a difference or a correlation coefficient \cite{robinson2001past}.

@@

Using the language of ``statistical clarity'' would help researchers escape the trap of accepting the null hypothesis, reduce multiple types of misinterpretation associated with accepting the null hypothesis, and allow for the continued use of NHST as a method of inference for for whether we can be confident of the sign of a difference (e.g. see \citealt{Abelson1997} for arguments for NHST). The use of ``significance'' to describe the results of hypothesis tests is deeply, and sometimes subtly, misleading, because it is at odds with other meanings of the word: the \pval is not an accurate gauge of whether a result is large in magnitude, biologically important, or relevant. ``Clarity,'' on the other hand, is an apt term for what NHST actually evaluates. \citet{jones2000sensible} and \citet{robinson2001past} suggest that researchers should report $p < 0.05$ using language such as ``the direction of the differences among the treatments was undetermined''. While this is a step in the right direction, replacing ``significance'' with ``clarity'' is a more complete, direct, and universal change that will improve communication and reduce sloppy thinking.

We often see sentences like, ``X et al. showed that there is no significant effect of Y on Z'' with the implication that this effect can now be assumed to be absent (or unimportant). In fact, the sentence is erroneous even before we get to the implication: X et al.'s significance test provides information about \emph{their study}, not about the underlying processes \citep{HoenigandHeisey2001}. Indeed, because a \pval is a property of the data, $p < 0.05$ can be obtained for a very small effect with a large enough sample size. A better formulation would be ``X et al. did not find a significant effect of Y on Z.'' This mistake is harder to make using the language of clarity:  ``X et al. showed that the effect of Y on Z is unclear'' is awkward. Similarly, ``We did not find a clear difference in response between the control and sham groups'' is both more colloquial and harder to transform into a misleading statement than ``We did not find a significant difference \ldots''. \citet{Bernardietal.2017} complained that ``\ldots sociological and social significance are sacrificed on the altar of statistical significance''. Describing statistical tests in terms of clarity would allow ``significant'' to reclaim its common English definition and reduce conflation between statistical results and substantive significance.

Descriptions of statistical results using the language of clarity should begin with reference to the effect. For example, ``The difference between the control and treatment group was not statistically clear.'' Table~\ref{quotetab} shows examples from the literature of language that misinterpret \pvals in three different ways, and demonstrates how to rewrite these statements in the language of clarity.

@@

Toxins accumulate after acute exposure but have no effect on behaviour
& Toxins accumulate after acute exposure but effects on behaviour are statistically unclear
\\

@@

\multicolumn{2}{c}{\emph{\textbf{Inferring weak effects from large \pvals}}}
\\
\hline
%% left ``P values'' unchanged here (not \pvals) since it's a quotation
\ldots\ differences between treatment and control groups were nonsignificant, with P values of at least 0.3, and most in the range $0.7 \leq P \leq 0.9$.
& \ldots\ differences between treatment and control groups were not statistically clear ($P > 0.05$) \ourcomment{confidence intervals would also be valuable here!}
\\ \\

@@

This correlation was significant in males ($\rho=0.35$, P \textless 0.05) but not females ($\rho=0.35$, NS). \ldots \ourcomment{The authors later assume that they have demonstrated a difference between males and females}
& Although males and females show the same correlation coefficient ($\rho=0.35$), the sign of the coefficient is statistically clear only in males  \ldots \ourcomment{This phrasing may suggest to the authors that confidence intervals are called for.}
\\
\rowcolor{lightgray}
\ldots risk of low BMD [bone mineral density] remained greater in HCV-coinfected women versus women with HIV alone
(adjusted OR 2.99, 95\% CI 1.33–6.74), but no association was found between HCV coinfection and low BMD in men 
(adjusted OR 1.26, 95\% CI 0.75–2.10). \ldots  \ourcomment{The authors do not directly compare men and women, but they state:}
The precise mechanisms for the association between viral hepatitis and low BMD in HIV-infected women but not men 
remain unclear.
& 
\ldots  risk of low BMD [bone mineral density] remained greater in HCV-coinfected women versus women with HIV alone
(adjusted OR 2.99, 95\% CI 1.33–6.74), but the association between HCV coinfection and low BMD in men was not
statistically clear (adjusted OR 1.26, 95\% CI 0.75–2.10). \ldots  Pursuing biological differences between women and men in the effect of HIV on BMD would be premature given these results.
\\
\end{tabular}
\captionof{table}{Examples of misinformed language in peer-reviewed papers (citations withheld out of courtesy, available by request) using our proposed  language of clarity.}
\label{quotetab}
\end{table}

\section*{Closing Comments}

@@

\bmb{would be good to have a citation here; do any of the ``best practices''/NHST criticisms bring up N-I tests? If not, is there a good intro reference?}
Following a shift in language, we suggest using a non-inferiority test to test for a small effect. Consider the following example. A hospital is considering switching from bulky, expensive surgical face masks to lighter weight, cheaper masks. A clinical trail that tests the effectiveness of both masks and finds $p > 0.05$ is not a success; this tells the hospital that the difference between the masks is not statistically clear! What the hospital actually want to know is whether the cheaper, lighter masks are \emph{only a little bit worse than the bulky, expensive masks}. This requires a $H_{0}$ that defines the target difference in effectiveness of the two masks that the hospital is willing to accept to make the transition. We note that this statistical test should be set up as a one-tailed test; the logical, \emph{a priori} assumption is that the bulky, expensive masks are more effective. In this case, a $p < 0.05$ tells the hospital that the difference between the new masks and old masks is small and that it they can proceed with the switch.

Accepting the null is especially common in diagnostic tests. For example, a Shapiro-Wilk test of normality that returns $p > 0.05$ means that it is statistically unclear how the data departs from normality (the data is collected from a complicated world and does depart from normality). Without the goal of picking on any paper in specific, ``\ldots allowing us to accept the null hypothesis about the normal spot count distribution.'' \citep{Karulinetal.2015} is common language that should be avoided.

@@

Following the lead of \citet{Cohen1994} and others \citep{Goodman1999, ZiliakandMcCloskey2008, WassersteinandLazar2016}, we conclude with a plea for an increased emphasis on effect sizes and confidence intervals. However, we suggest that the use of ``statistical clarity'' forces the \emph{need} for confidence intervals and effect sizes by removing ambiguity from statements associated with $p < 0.05$. The statement ``The difference between our control and treatment groups was statistically clear ($p = 0.007$)'' is unambiguously incomplete; an effect size and confidence interval is required to complete the story.

We echo previous statements in favor of ``neglected factors'' (prior and related evidence, plausibility of mechanism, study design and data quality, real world benefits, novelty and other factors) \citep{McShaneetal.2017} and reporting of \emph{a priori} analysis of statistical power to avoid emphasis on implausibly large effects given low statistical power \citep[the ``winner's curse''][]{GelmanandCarlin2014, SzucsandIoannidis2017, Bernardietal.2017}.  Additionally, we support the idea of writing a statistical journal that chronicles all steps in the analysis process \citep{Kassetal.2016}, and clearly defining the boundary between results presented as inferences based on \emph{a priori} hypotheses and those from \emph{post hoc} data exploration.

>>>>

  \textbf{Abstract}: Null hypothesis significance testing remains popular despite decades of concern about misuse and misinterpretation. We believe that a significant part of the problem is due simply to language: significance testing has little to do with other meanings of the word ``significance''. Despite the limitations of null-hypothesis tests, we argue here that they remain useful in many contexts as a guide to whether a certain effect can be seen \emph{clearly} in that context (e.g. whether we can clearly see that a correlation or between-group difference is positive or negative). We therefore suggest that researchers describe the conclusions of null-hypothesis tests in terms of statistical ``clarity'' rather than statistical ``significance''. This simple semantic change could substantially enhance clarity in statistical communication.

@@

Statisticians and scientists have bemoaned the shortcomings of null hypothesis significance testing (NHST) for nearly a century \citep{Cohen1994}.
Books and articles proposing the de-emphasis or abandonment of the \pval have been cited thousands of times \citep{Cohen1994, Goodman1999, Wilkinson1999, ZiliakandMcCloskey2008, WassersteinandLazar2016}. These works plead for a focus on effect sizes and confidence intervals, and point out that null effects that truly have zero magnitude are unrealistic or impossible in most fields outside of the hard physical sciences \citep{Meehl1990, Tukey1991, Cohen1994}. Yet, \pvals without confidence intervals (or even effect sizes) and references to null effects still pervade the scientific literature at all levels up to and including articles in high-impact journals.

In a meta-analysis of 356 studies \citet{Bernardietal.2017} found that 72\% of studies contained an ambiguous use of the term ``significant'', 49\% interpreted non-significant effects as zero effects, and 44\% failed to report a comprehensible effect size. The misuse and misinterpretation of NHST is so frequent that there have been recent calls for drastically reducing \citep{SzucsandIoannidis2017} or abandoning \citep{McShaneetal.2017} its use. Other prescriptions have included the complete abandonment of frequentist statistics \bmb{CITE: is it OK to include this here, or it too distracting?}, or the use of a stricter significance threshold (e.g. $p < 0.005$: \citealt{Benjaminetal.2018}); however, the former seems impractical, while the latter is unlikely to reduce the misuse and misinterpretation of \pvals, or the publication bias imposed by any \pval threshold \citep{Ridleyetal.2007}.

\citet{Kramer2011} points out Ziliak and McCloskey's (\citeyear{ZiliakandMcCloskey2008}) query how could NHST possibly have survived ``\ldots if it is as idiotic as we and other critics have so long believed''. We believe that NHST itself \bmb{[although dangerously prone to misuse?]} is not as idiotic as suggested; the idiocy associated with NHST is the language we use in describing it. A strong argument in favor of NHST is that testing the null hypothesis (even though we usually know that it's false, and can never prove that it's true) is a proxy for asking whether we can see a difference clearly. In many cases, this comes down simply to whether we can be confident of the \emph{sign} of a difference or a correlation coefficient \citep{robinson2001past}.

@@

Using the language of ``statistical clarity'' would help researchers escape the trap of accepting the null hypothesis, reduce multiple types of misinterpretation associated with NHST, and allow for the continued use of NHST to determine whether we can be confident about the sign of a difference (e.g. see \citealt{Abelson1997} for arguments for NHST \bmb{what's in Abelson? How much of our argument do they anticipate?}). The use of ``significance'' to describe the results of hypothesis tests is deeply, and sometimes subtly, misleading, because it is at odds with other meanings of the word: the \pval is not an accurate gauge of whether a result is large in magnitude, biologically important, or relevant. ``Clarity,'' on the other hand, is an apt term for what NHST actually evaluates. \citet{jones2000sensible} and \citet{robinson2001past} suggest that researchers should report $p < 0.05$ using language such as ``the direction of the differences among the treatments was undetermined''. While this is a step in the right direction, replacing ``significance'' with ``clarity'' is a more complete, direct, and universal change that will improve communication and reduce sloppy thinking.

We often see sentences like, ``X et al. showed that there is no significant effect of Y on Z'' with the implication that this effect can now be assumed to be absent (or unimportant). In fact, the sentence is erroneous even before we get to the implication: X et al.'s significance test provides information about \emph{their particular study}, not about the underlying processes \citep{HoenigandHeisey2001}. Indeed, because a \pval is a property of the data, $p < 0.05$ can be obtained for a very small effect with a sufficently large  sample, or for a very large effect with a sufficiently small or noisy sample. A better formulation would be ``X et al. did not find a significant effect of Y on Z.'' This mistake is harder to make using the language of clarity:  ``X et al. showed that the effect of Y on Z is unclear'' is awkward. Similarly, ``We did not find a clear difference in response between the control and sham groups'' is both more colloquial and harder to transform into a misleading statement than ``We did not find a significant difference \ldots''. \citet{Bernardietal.2017} complained that ``\ldots sociological and social significance are sacrificed on the altar of statistical significance''. Describing statistical tests in terms of clarity would allow ``significant'' to reclaim its common English definition and reduce conflation between statistical results and substantive significance.

Descriptions of statistical results using the language of clarity should begin with reference to the effect. For example, ``The difference between the control and treatment group was not statistically clear.'' Table~\ref{quotetab} shows published examples of statements that misinterpret \pvals in three different ways and demonstrates how to rephrase them in the language of clarity.

@@

Toxins accumulate after acute exposure but have no effect on behaviour
& Toxins accumulate after acute exposure but their effects on behaviour are statistically unclear
\\

@@

\multicolumn{2}{c}{\emph{\textbf{Inferring weak effects from large \pvals}} \citep{WassersteinandLazar2016}}
\\
\hline
%% left ``P values'' unchanged here (not \pvals) since it's a quotation
\ldots\ differences between treatment and control groups were nonsignificant, with P values of at least 0.3, and most in the range $0.7 \leq P \leq 0.9$.
& \ldots\ differences between treatment and control groups were not statistically clear (all $P > 0.05$) \ourcomment{confidence intervals would also be valuable here!}
\\ \\

@@

This correlation was significant in males ($\rho=0.35$, P \textless 0.05) but not females ($\rho=0.35$, NS). \ldots \ourcomment{The authors later write as though they have demonstrated a difference between males and females}
& Although males and females show the same correlation coefficient ($\rho=0.35$), the sign of the coefficient is statistically clear only in males  \ldots \ourcomment{This phrasing may suggest to the authors that confidence intervals are called for.}
\\
\rowcolor{lightgray}
\ldots risk of low BMD [bone mineral density] remained greater in HCV-coinfected women versus women with HIV alone
(adjusted OR 2.99, 95\% CI 1.33–6.74), but no association was found between HCV coinfection and low BMD in men 
(adjusted OR 1.26, 95\% CI 0.75–2.10). \ldots  \ourcomment{The authors do not directly compare men and women, but they state:}
The precise mechanisms for the association between viral hepatitis and low BMD in HIV-infected women but not men 
remain unclear.
& 
\ldots  risk of low BMD [bone mineral density] remained greater in HCV-coinfected women versus women with HIV alone
(adjusted OR 2.99, 95\% CI 1.33–6.74), but the association between HCV coinfection and low BMD in men was not
statistically clear (adjusted OR 1.26, 95\% CI 0.75–2.10). \ldots  Pursuing biological differences between women and men in the effect of HIV on BMD would be premature given these results.
\\
\end{tabular}
\captionof{table}{Examples of misinformed language in peer-reviewed papers (citations withheld out of courtesy, available by request) using our proposed  language of clarity. \bmb{do we want to comment on our search strategy? Doesn't need to be formal [this isn't a meta-analysis], but might be of interest to readers}}
\label{quotetab}
\end{table}

\section*{Further recommendations}

\bmb{this section seems a bit ``grab-baggy''. Can we restructure it so it feels more coherent? I don't mind including these points, but I would rather leave them out than dilute the message/bog the reader down. Do we want to mention the more difficult case of NHST on non-negative domains (i.e. joint tests of multiple parameters, tests of variances, etc., where we can't interpret \pval as clarity about signs?)}

\bmb{I'd like to consider omitting the next (sub)section. The connection between the title and the text is not very clear, and I don't think it's the most important point. (If it can be made clear and punchy I don't mind including it.) Interesting discussion of using Shapiro-Wilk $W$ as an effect size (spoiler: it might be a bad idea)  \underline{\href{https://stats.stackexchange.com/questions/2492/is-normality-testing-essentially-useless}{here}}, and general discussion of uselessness of Normality testing \underline{\href{https://stats.stackexchange.com/questions/289136/is-the-shapiro-wilk-test-w-an-effect-size}{here}}. Also, we should be able to invent some  typically bad wording without breaking our protocol and mentioning a particular article here: ``Wording such as `The Shapiro-Wilk test gave $p>0.05$, allowing us to conclude that the data are normally distributed' \ldots''}

Accepting the null is especially common when researchers perform diagnostic tests of the assumptions of statistical tests. For example, a Shapiro-Wilk test of normality that returns $p > 0.05$ means that it is statistically unclear how the data departs from normality (the data is collected from a complicated world and does depart from normality). Without the goal of picking on any paper in specific, ``\ldots allowing us to accept the null hypothesis about the normal spot count distribution.'' \citep{Karulinetal.2015} is common language that should be avoided.

@@

\bmb{(1) would be good to have a citation here; do any of the ``best practices''/NHST criticisms bring up N-I tests? If not, is there a good intro reference? If it's \emph{not} widely stated in the NHST-criticism literature, that strengthens the case for including it here. (2) This needs more context, if we include it; if as suggested it's not widely pointed out in the NHST-crit lit, ``One tool that is well known in the biostatistical world but underappreciated in more applied areas \ldots''. Or, ``Reporting CIs provides meaningful information about how small a measured effect might be \ldots if researchers want to be perform a formal test that biological effects are small \ldots'' As I understand it non-inferiority tests are directed; could one construct a p-value for an \emph{unsigned} test, i.e. trying to show that $|\beta|<T$ for some pre-specified $T$? Or am I sneaking up on an \emph{a priori} power analysis?}

Following a shift in language, we suggest using a non-inferiority test to test for the hypothesis that biological effects are small. Consider the following example. A hospital is considering switching from bulky, expensive surgical face masks to lighter weight, cheaper masks. A clinical trail that tests the effectiveness of both masks and finds $p > 0.05$ is not a success; this tells the hospital that the difference between the masks is not statistically clear! What the hospital actually want to know is whether the cheaper, lighter masks are \emph{only a little bit worse than the bulky, expensive masks}. This requires a $H_{0}$ that defines the target difference in effectiveness of the two masks that the hospital is willing to accept to make the transition. We note that this statistical test should be set up as a one-tailed test; the logical, \emph{a priori} assumption is that the bulky, expensive masks are more effective. In this case, a $p < 0.05$ tells the hospital that the difference between the new masks and old masks is small and that it they can proceed with the switch.

@@

\bmb{I'd be OK with making this section alone be our Conclusions (if we don't want to clean up/improve the previous sections)}
Following the lead of \citet{Cohen1994} and others \citep{Goodman1999, ZiliakandMcCloskey2008, WassersteinandLazar2016}, we conclude with a plea for an increased emphasis on effect sizes and confidence intervals. However, we suggest that the use of ``statistical clarity'' reinforces the need for confidence intervals and effect sizes by making it clearer that bald statements about \pvals are insufficient. The statement ``The difference between our control and treatment groups was statistically clear ($p = 0.007$)'' is unambiguously incomplete; an effect size and confidence interval are required to complete the story.

We echo previous statements in favor of ``neglected factors'' (prior and related evidence, plausibility of mechanism, study design and data quality, real world benefits, novelty and other factors) \citep{McShaneetal.2017} and reporting of \emph{a priori} analysis of statistical power to avoid emphasis on implausibly large effects given low statistical power \citep[the ``winner's curse''][]{GelmanandCarlin2014, SzucsandIoannidis2017, Bernardietal.2017}.  Additionally, we support the idea of writing a statistical journal that chronicles all steps in the analytical process \citep{Kassetal.2016}, and clearly delineating the boundary between inferences based on \emph{a priori} hypotheses and discoveries from \emph{post hoc} data exploration.

Whether or not our recommendations are broadly adopted by authors, reviewers, and editors, they can be useful for individual researchers who want to help themselves think clearly about NHST results. We have found that rephrasing NHST statements that we encounter (in the literature, or in seminar presentations) in terms of clarity has helped us to correctly interpret their meanings.
